dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format

### 2022-11-01 ms.Qmd
Fix yaml header

```
title: 'dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format'
tags:
  - R
  - taxonomy
  - biodiversity
authors:
  - name: Joel H. Nitta
    orcid: 0000-0003-4719-7472
    affiliation: 1
  - name: Wataru Iwasaki
    orcid: 0000-0002-9169-9245
    affiliation: 1
affiliations:
  - name: The University of Tokyo, Japan
date: 31 October 2022
bibliography: references.bib
```

@@

The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biological data.
`dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
`dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.

@@

Taxonomic data fundamentally represent scientific hypotheses [@Sluys2013].
As such, taxonomic data are continuously being updated as new species are discovered, old taxonomic concepts are revised, etc.
Such taxonomic changes can be complicated.
For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
Because of these properties, maintaining a taxonomic database in compliance with DWC can be onerous and error prone if done by hand.
There is therefore a need for software to automate entry and validation of DWC taxonomic data.

@@

Validation checks in `dwctaxon` include the following (each is a logical argument of `dct_validate()`):
- `check_taxon_id`: Check that all values of `taxonID` are unique and non-missing.
- `check_tax_status`: Check that all values of `taxonomicStatus` are valid. Default valid values include "accepted", "synonym", or "variant", but these can be set via the `valid_tax_status` argument.
- `check_mapping`: Check that all values of `acceptedNameUsageID` map to the `taxonID` of an existing name.
- `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).
- `check_sci_name`: Check that all instances of `scientificName` are non-missing and unique.
- `check_status_diff`: Check that each scientific name be allowed to have only one taxonomic status (typically only used if `check_sci_name` is not used).
- `check_col_names`: Check that all column names are valid Darwin Core terms.

The are multiple options for the output of `dct_validate()`:
- If the `on_fail` argument is set to `error`, an error will be raised and computation stopped at the first failing check.
- If `on_fail` is set to `summary`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.
- If the `on_success` argument is set to `data` and all checks pass, the original data will be returned.
- If `on_success` is set to `logical` and all checks pass, a logical value of `TRUE` will be returned.

@@

## Example: GBIF

@@

@@

The only similar tool we are aware of is the GBIF data validator, an online system for validating data in DWC format.
However, the GBIF data validator is not available for use as a local program and is limited to input sizes of 100 mb.
Furthermore, it does not conduct editing of DWC compliant data.

>>>>

\-\-\-
title: dwctaxon, an R package for editing and validating taxonomic data in Darwin Core format
tags:
\ \ - R
\ \ - taxonomy
\ \ - biodiversity
authors:
\ \ - name: Joel H. Nitta
\ \ \ \ orcid: 0000-0003-4719-7472
\ \ \ \ affiliation: 1
\ \ - name: Wataru Iwasaki
\ \ \ \ orcid: 0000-0002-9169-9245
\ \ \ \ affiliation: 1
affiliations:
\ \ - name: The University of Tokyo, Japan
date: 31 October 2022
bibliography: references.bib
\-\-\-

@@

The `dwctaxon` R package provides functions to edit and validate taxonomic data in compliance with the widely used Darwin Core (DWC) standard [@Wieczorek2012] for biodiversity data.
`dwctaxon` automates several data editing steps, thereby simplifying workflows and decreasing the chance of human-generated data entry errors.
Furthermore, it conducts automated checks to validate taxonomic datasets, thereby alerting database maintainers to problems as soon as possible.
`dwctaxon` will accelerate the generation and maintenance of taxonomic data, a critical part of the effort to document and conserve earth's biodiversity.

@@

Taxa fundamentally represent scientific hypotheses [@Sluys2013].
As such, taxonomic data are continuously being updated as new species are discovered, old taxonomic concepts are revised, etc.
Such taxonomic changes can be complicated.
For example, changing the status of one name to the synonym of another could potentially involve modifying many other names if they are synonyms of the first.
Because of these properties, maintaining a taxonomic database in compliance with DWC can be onerous and error prone if done by hand.
There is therefore a need for software to automate entry and validation of DWC taxonomic data.

@@

Validation checks in `dwctaxon` include the following (each is a logical argument of `dct_validate()`):

- `check_taxon_id`: Check that all values of `taxonID` are unique and non-missing.

- `check_tax_status`: Check that all values of `taxonomicStatus` are valid. Default valid values include "accepted", "synonym", or 
"variant", but these can be set via the `valid_tax_status` argument.

- `check_mapping`: Check that all values of `acceptedNameUsageID` map to the `taxonID` of an existing name.

- `check_mapping_strict`: Check that mapping of names follow expectations based on their taxonomic status (e.g., synonyms must map to accepted names, accepted names cannot have an `acceptedNameUsageID`, etc.).

- `check_sci_name`: Check that all instances of `scientificName` are non-missing and unique.

- `check_status_diff`: Check that each scientific name be allowed to have only one taxonomic status (typically only used if `check_sci_name` is not used).

- `check_col_names`: Check that all column names are valid Darwin Core terms.

The are multiple options for the output of `dct_validate()`:

- If the `on_fail` argument is set to `error`, an error will be raised and computation stopped at the first failing check.

- If `on_fail` is set to `summary`, if any check fails a summary table (tibble) will be returning describing failures for all checks, with a warning.

- If the `on_success` argument is set to `data` and all checks pass, the original data will be returned.

- If `on_success` is set to `logical` and all checks pass, a logical value of `TRUE` will be returned.

@@

## Example usage

@@

## Caveats

@@

The only similar tool we are aware of is the GBIF data validator (<https://www.gbif.org/tools/data-validator>), an online system for validating data in DWC format.
However, the GBIF data validator is not available for use as a local program and is limited to input sizes of 100 mb.
It does not conduct editing of DWC compliant data, nor are we aware of any other software that does so.
